---
title: "02: Cohort Selection and LOCF Imputation"
author: "Anel Zhunussova and Poornima Kumar"
date: "9 February 2026"
output:
  html_document:
    df_print: paged
    toc: true
    theme: united
    number_sections: false
editor_options: 
  markdown: 
    wrap: 72
---

# Script overview

**Purpose:**

This script serves as the cohort selection and data filtering phase. It transforms the longitudinal "master_database" (created in script 01) into a final cross-sectional dataset ready for modeling. Mainly, the script works on selecting scales based on the earliest available vaccination intent response for each participant. To maximize data retention, it utilizes a Last Observation Carried Forward (LOCF) approach, pairing the vaccination intent outcome with the most recent available predictor data (either concurrent or prior) to ensure valid responses are not discarded due to missing concurrent scales.

**Requirements:**
This script requires the following file in the `data/processed/` directory:
- `master_database.csv`: Generated in Script 01.

**Outline:**
1. **Setup and preparation:** Loading centralized packages, global configurations, and the master database.
2. **Data selection:** Reshaping to long format, applying 6-week LOCF imputation, and selecting the index time-point (earliest response).
3. **Final cleanup:** Formatting variables, filtering by recruitment density (Months with N >= 30), and removing incomplete cases.
4. **Supplementary QC:** Validating the timeline distribution, temporal bias in intent, and LOCF imputation burden/lag.

**Output:** 
- `cohort_data.csv`: Saved in `data/processed/`.
- Figures for supplementary results.

---

# 1. Setup and preparation

Loading project utilities and the master database.
```{r setup, message=FALSE, warning=FALSE}
# setwd("../scripts") # Adjust path if running manually

source("../R/load_packages.R")
source("../R/config.R")
source("../R/processing_utils.R") 
source("../R/visualization_utils.R")

# Loading the master database from the processed data folder
master_data <- read.csv("../data/processed/master_database.csv")
```

# 2. Data selection
## 2.1 Reshaping to long format
Pivoting time-varying columns to rows while keeping static demographic variables constant. We use the target timepoints and static variables defined in config.R.

```{r}
long_data <- master_data %>% 
  pivot_longer(
    cols = -all_of(static_vars), 
    names_to = c(".value", "timepoint"), 
    names_pattern = "(.*)_(.*)$")  %>%
  
  # Factoring timepoints using levels from config
  mutate(timepoint = factor(timepoint, levels = target_timepoints, ordered = TRUE)) %>%
  arrange(record_id, timepoint)
```


## 2.2 Imputation (LOCF)
Filling missing predictor values using the Last Observation Carried Forward method. We implement a 6-week look-back window (checking concurrent, then up to 3 prior bi-weekly steps).

```{r}
# Defining time-varying predictor columns for imputation
predictor_cols <- setdiff(names(long_data), c(static_vars, "timepoint", "take_vaccination", "date_complete"))

long_filled <- long_data %>%
  mutate(date_complete = lubridate::as_date(date_complete)) %>% 
  group_by(record_id) %>%
  arrange(timepoint, .by_group = TRUE) %>%
  
  # Recovering missing dates within the 6-week window
  mutate( 
    date_was_recovered = is.na(date_complete),
    date_complete = case_when(
      !is.na(date_complete)        ~ date_complete,
      !is.na(lag(date_complete, 1)) ~ lag(date_complete, 1) + 14,
      !is.na(lag(date_complete, 2)) ~ lag(date_complete, 2) + 28,
      !is.na(lag(date_complete, 3)) ~ lag(date_complete, 3) + 42,
      TRUE                          ~ as.Date(NA) 
    ),
  
  # Implementing 6-week look-back for predictors
  across(all_of(predictor_cols), 
                ~ case_when(
                  !is.na(.)        ~ .,         
                  !is.na(lag(., 1)) ~ lag(., 1), 
                  !is.na(lag(., 2)) ~ lag(., 2), 
                  !is.na(lag(., 3)) ~ lag(., 3), 
                  TRUE              ~ NA_real_  
                ))) %>%
  ungroup() 

# Selecting the earliest available vaccine response per participant
selected_data <- long_filled %>%
  filter(!is.na(take_vaccination)) %>%
  group_by(record_id) %>%
  slice_min(order_by = timepoint, n = 1, with_ties = FALSE) %>%
  ungroup()
```

# 3. Final cleanup
## 3.1 Cleaning 
Ensuring scale variables are numeric and removing incomplete cases. We also filter for high-density months to reduce temporal bias.
```{r}
# Initial cleanup and date formatting
cohort_data <- selected_data %>%
  mutate(across(-c(record_id, timepoint, date_complete, sex, area_type), as.numeric)) %>%
  drop_na() %>%
  mutate(month = floor_date(as_date(date_complete), "month"))

# Identifying months with at least 30 participants
months_to_keep <- cohort_data %>%
  group_by(month) %>%
  summarise(n = n(), .groups = "drop") %>%
  filter(n >= 30) %>%
  pull(month)

# Final filtering and applying labels from config
cohort_data_final <- cohort_data %>%
  filter(month %in% months_to_keep) %>%
  mutate(take_vaccination_lbl = apply_factor(take_vaccination, "take_vaccination"))

cat(sprintf("Final Model Data N: %d\nFiltered months: %d\n", 
            nrow(cohort_data_final), length(months_to_keep)))
```

## 3.2 Saving the finalized cohort data
```{r}
#write.csv(cohort_data_final, "../data/processed/cohort_data.csv", row.names = FALSE) # 962 obs 65 vars (includes intermittent QC vars)
```

```{r}
# Summary statistics (more in 03 scripts)
cat("======================================================\n")
cat(" FINAL MODEL DATA \n")
cat("======================================================\n")
cat(sprintf("Total Participants (N): %d\n", nrow(cohort_data_final)))
cat(sprintf("Data Collection Period: %s to %s\n", min(cohort_data_final$date_complete, na.rm = TRUE), max(cohort_data_final$date_complete, na.rm = TRUE)))
cat("------------------------------------------------------\n")
cat("Timepoint Distribution:\n")
print(table(cohort_data_final$timepoint))
cat("======================================================\n")
```

# 4. Supplementary: Quality control
## 4.1 Vaccine response exploration
Visualizing the data collection timeline and the temporal distribution of intent.
```{r}
p_timeline <- plot_timeline(cohort_data_final)
p_bias     <- plot_outcome_temporal_bias(cohort_data_final)

p_timeline
p_bias
```

## 4.2 LOCF validation
Calculating the temporal lag for each imputed observation to validate the 6-week window.
```{r}
# Calculating the lag per scale for each participant
time_ranks <- data.frame(timepoint = target_timepoints, rank_id = seq_along(target_timepoints))

targets <- cohort_data_final %>%
  select(record_id, target_timepoint = timepoint) %>%
  left_join(time_ranks, by = c("target_timepoint" = "timepoint")) %>%
  rename(target_rank = rank_id)

sources <- long_data %>%
  filter(record_id %in% cohort_data_final$record_id) %>%
  select(record_id, timepoint, all_of(predictor_cols)) %>%
  pivot_longer(cols = all_of(predictor_cols), names_to = "variable", values_to = "value") %>%
  filter(!is.na(value)) %>%
  left_join(time_ranks, by = "timepoint") %>%
  rename(source_rank = rank_id)

lag_table <- sources %>%
  inner_join(targets, by = "record_id") %>%
  filter(source_rank <= target_rank) %>%
  group_by(record_id, variable) %>%
  summarise(lag_steps = first(target_rank) - max(source_rank), .groups = "drop")

```

Getting LOCF summary 
```{r}
LOCF_summary <- prep_imputation_summaries(lag_table) 
```

Plotting imputation results
```{r fig.width=9.5, fig.height=9}
p_incidence <- plot_imputation_incidence(LOCF_summary$incidence_plot_data)
p_burden    <- plot_imputation_burden(LOCF_summary$imputed_only_plot_data)
p_lag       <- plot_lag_by_scale(LOCF_summary$locf_plot_data, LOCF_summary$sort_order)

p_incidence
p_burden
p_lag
```

# 5 Saving figures 
```{r}
# Creating a list of all QC figures for Script 02
s02_qc_plots <- list(
  (p_timeline + p_bias) / (p_incidence + p_burden),
  p_lag
)

# Saving as PDF 
#save_report_pdf(s02_qc_plots, "../figures/QC_Cohort_Selection_LOCF.pdf",  width = 13, height = 12)
```

