---
title: "03: Exploratory Data Analysis"
author: "Anel Zhunussova and Poornima Kumar"
date: "9 February 2026"
output:
  html_document:
    df_print: paged
    toc: true
    theme: united
    number_sections: false
editor_options: 
  markdown: 
    wrap: 72
---

# Script overview

**Purpose:**
This script serves as QC and data exploration for the cohort data. We visualize the distributions of sociodemographic and psychological predictors, calculate group differences via Cohen's D, and assess multicollinearity through correlation analysis.

**Requirements:**
- `cohort_data.csv`: Created in Script 02.

**Outline:**
1. **Setup and preparation:** Loading utilities and factoring data.
2. **Sociodemographic factors:** Distributions.
3. **Pandemic context:** Distributions.
3. **Psychological profiles:** Raincloud plots (with Bayesian t-tests) and ridge plots.
4. **Overall analysis:** Correlation matrices and Cohen’s D effect sizes.

---

# 1. Setup and preparation

Loading project utilities and the cohort dataset.
```{r setup, message=FALSE, warning=FALSE}
# setwd("../scripts")

source("../R/load_packages.R")
source("../R/config.R")
source("../R/processing_utils.R")
source("../R/visualization_utils.R")

# Load data and apply factor mapping from config
cohort_data <- read.csv("../data/processed/cohort_data.csv") %>%
  mutate(
    # Convert the pre-existing label to a factor to ensure plot order
    take_vaccination_lbl = factor(take_vaccination_lbl, 
                                  levels = c("Intention (Yes)", "Hesitancy (No)")),
    
    # Automatically factorize everything else that has a key in levels_map
    across(any_of(names(levels_map)), ~ apply_factor(.x, cur_column()))
  )
```

# 2. Sociodemographic predictors
Visualizing the distribution of demographic parameters across vaccination intent.
```{r fig.width=15, fig.height=9}
p_demo_total <- plot_var_grid(cohort_data, demo_vars, ncol = 5, title = "Demographics")
p_demo_total
```

# 3. Pandemic context predictors
Visualizing variables related to pandemic exposure, distancing, and health impact.
```{r fig.width=22, fig.height=11}
p_context_total <- plot_var_grid(cohort_data, context_vars, ncol = 6, title = "Pandemic Context") 
p_context_total
```

# 4. Psychological predictors
## 4.1 Raincloud
Comparing psychological scales between intenders and refusers using raincloud plots.
```{r fig.width=15, fig.height=5}
# Defining the recipe for all psychological scales
plot_psych_recipe <- list(
  promis = list(prefix = "promis_", title = "PROMIS",
    vars = colnames(cohort_data %>% select(starts_with("promis_")))),
  apathy = list(prefix = "apathy_", title = "APATHY",
    vars = colnames(cohort_data %>% select(starts_with("apathy_")))),
  coping1 = list(prefix = "coping_", title = "COPING (A)",
    vars = c("coping_accept", "coping_beh_diseng", "coping_denial", "coping_emo_support", 
             "coping_religion", "coping_active","coping_humor")),
  coping2 = list(prefix = "coping_", title = "COPING (B)",
    vars = c("coping_inst_support", "coping_plan", "coping_pos_fram", "coping_self_blame", 
             "coping_self_dist", "coping_substance", "coping_venting")),
  other = list(prefix = "", title = "Other bi-weekly",
    vars = c("loneliness", "anhedonia")),
  bis = list(prefix = "bis_", title = "BIS",
    vars = colnames(cohort_data %>% select(starts_with("bis_")))),
  mpss = list(prefix = "mpss_", title = "MPSS",
    vars = colnames(cohort_data %>% select(starts_with("mpss_")))),
  brs = list(prefix = "", title = "BRS", vars = "brs"),
  pss = list(prefix = "", title = "PSS", vars = "pss"),
  other_monthly = list(prefix = "", title = "Other (monthly)",
    vars = c("sias", "worry"))
)

# Generating plots with Bayesian t-tests
current_test <- "bayesian" 

all_psych_figs <- map(plot_psych_recipe, ~{
  plot_rainclouds(
    data = cohort_data,
    group_var = "take_vaccination_lbl",
    metric_vars = .x$vars,
    prefix_to_remove = .x$prefix,
    title = .x$title,
    test_type = current_test 
  )})
```

## 4.2 Ridge plots
```{r}
p_ridge_biweekly <- (
  make_ridge_plot(cohort_data, plot_psych_recipe$promis$vars, "PROMIS") | 
  make_ridge_plot(cohort_data, plot_psych_recipe$apathy$vars, "APATHY")
) / make_ridge_plot(cohort_data, plot_psych_recipe$coping1$vars, "COPING") +
  plot_layout(guides = "collect") & theme(legend.position = "bottom")

p_ridge_monthly <- (
  make_ridge_plot(cohort_data, plot_psych_recipe$bis$vars, "BIS") |
  make_ridge_plot(cohort_data, plot_psych_recipe$mpss$vars, "MPSS")
) / (
  make_ridge_plot(cohort_data, c("sias", "worry"), "Other") |
  make_ridge_plot(cohort_data, c("pss", "brs"), "Stress & Resilience")
) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

# 5. Overall analysis
## 5.1 Correlation
Assessing Spearman correlations to identify multicollinearity within predictor blocks and for all.
```{r fig.width=10, fig.height=10}
# Correlation for all variables
all_vars <- c(demo_vars, context_vars, psych_vars)
p_cor_all <- plot_category_cor(cohort_data, all_vars, "Correlation Matrix: All Predictors & Outcome")

# Correlations by category
p_cor_demo <- plot_category_cor(cohort_data, demo_vars, "Sociodemographic & Outcome")
p_cor_ctx  <- plot_category_cor(cohort_data, context_vars, "Pandemic Context & Outcome")
p_cor_psyc <- plot_category_cor(cohort_data, psych_vars, "Psychological Scales & Outcome")
```

## 5.2 Dendogram
Dendogram to check for redundant predictors
```{r fig.width=12, fig.height=14}
p_dendro <- plot_dendrogram(cohort_data, all_vars)
p_dendro
```

## 5.3 Cohen's D
Calculating the standardized effect size for each predictor.
```{r fig.width=12, fig.height=14}
p_cohen <- plot_cohens_d_summary(cohort_data, all_vars, "take_vaccination_lbl")
p_cohen
```

# 6 Saving figures
```{r}
s03_report_plots <- list(
  p_demo_total, 
  p_context_total,

  all_psych_figs$promis / (all_psych_figs$apathy + all_psych_figs$other),
  all_psych_figs$coping1 / all_psych_figs$coping2 / all_psych_figs$bis,
  (all_psych_figs$mpss + all_psych_figs$brs) / (all_psych_figs$pss + all_psych_figs$other_monthly),
  
  p_ridge_biweekly,
  p_ridge_monthly,
  
  p_cor_all,
  p_cor_psyc,
  p_cor_demo + p_cor_ctx,
  p_dendro,
  
  p_cohen
)
#save_report_pdf(s03_report_plots, "../figures/QC_EDA.pdf", width = 20, height = 12)
```

