---
title: "04: Feature engineering and Data Splitting"
author: "Anel Zhunussova and Poornima Kumar"
date: "9 February 2026"
output:
  html_document:
    df_print: paged
    toc: true
    theme: united
---

# Script overview

**Purpose:**
This script prepares the dataset for machine learning modeling. We perform final feature cleaning (applying correct data types and removing metadata) and execute a stratified train/test split (80/20). Stratification ensures that both the training and hold-out sets are representative of the original population's vaccine intent, sex, and age distribution.

**Requirements:**
- `cohort_data.csv`: Created in Script 02.

**Outline:**
1. **Setup and preparation:** Loading utilities and cohort data.
2. **Data splitting:** Creating the stratified partition index.
3. **Split Quality Control:** Visualizing the balance between training and testing sets.
4. **Cleaning and Formatting:** Feature cleaning.
4. **Data Export:** Saving the final partitioned data as an RDS object.

---

# 1. Setup and preparation
```{r setup, message=FALSE, warning=FALSE}
source("../R/load_packages.R")
source("../R/config.R")
source("../R/modeling_utils.R")    
source("../R/visualization_utils.R")

cohort_data <- read.csv("../data/processed/cohort_data.csv")

# Just for safety:
# Sort by record_id to ensure that "Row 1" is always the same human being.
# This ensures that set.seed(123) produces the exact same split every time.
cohort_data <- cohort_data %>% arrange(record_id)
```

# 2. Data Splitting
We perform a stratified 80/20 split. To prevent bias and ensure the test set is representative, we stratify by Vaccination Intent, Sex, and Age (binned into quartiles).
```{r}
set.seed(123) 

# Create strata based on Outcome + Sex + Binned Age
age_bins <- cut(cohort_data$age, breaks = 4, labels = FALSE)
strata_vector <- paste(cohort_data$take_vaccination, cohort_data$sex, age_bins, sep = "_")

# Generate 80/20 split index
train_index <- createDataPartition(strata_vector, p = 0.80, list = FALSE)
```

# 3. Split Quality Control
Visualizing if the Training and Testing sets are balanced.
```{r}
# Prepare dataframe for QC plotting
plot_df <- cohort_data %>%
  mutate(split = if_else(row_number() %in% train_index, "Train", "Test"),
         split = factor(split, levels = c("Train", "Test")))

p_age_qc  <- plot_split_compare(plot_df, "age", type = "continuous")
p_sex_qc  <- plot_split_compare(plot_df, "sex", type = "categorical")
p_out_qc  <- plot_split_compare(plot_df, "take_vaccination", type = "categorical")

fig_split_qc <- (p_age_qc + p_sex_qc + p_out_qc) +
  plot_layout(guides = "collect") &
  plot_annotation(title = "Stratified Split QC: Train vs Test Balance") &
  theme(legend.position = "bottom", legend.title = element_blank())

fig_split_qc

#save_report_pdf(list(fig_split_qc), "../figures/QC_Data_Split.pdf", width = 12, height = 6)
```
# 4. Cleaning & Formating 
```{r}
# We convert variables to their correct types (Factors/Numeric) and remove metadata, dates to ensure the model only learns from raw predictors.
processed_df <- prepare_model_data_base(
  df = cohort_data, 
  target_var = "take_vaccination", 
  nominal_list = nominal_vars
)
```

# 5. Scenario Creation and Export
We save the data as a single list object named model_split. This object contains the distinct Training and Testing sets, ensuring consistency across all subsequent modeling scripts.
```{r}
# Construct the split object
model_split <- list(
  train = processed_df[train_index, ],
  test  = processed_df[-train_index, ]
)

# Save to the processed data directory
#saveRDS(model_split, "../data/final/model_split.rds")

# Final confirmation for report
message(">>> Data successfully split.")
message(">>> Training N: ", nrow(model_split$train))
message(">>> Testing N: ",  nrow(model_split$test))
```


