---
title: '06: Final Model and Interpretation'
author: "Anel Zhunussova and Poornima Kumar"
date: "9 February 2026"
output:
  pdf_document:
    toc: true
  html_document:
    df_print: paged
    toc: true
    theme: united
---

# Script overview
**Purpose:** This script retrains our selected Elastic Net model on the full training set and validates it against the hold-out set. We perform a deep-dive interpretation using SHAP values, discover hierarchical interactions, and identify "Phenotype Profiles" via clustering to explain the drivers of vaccine hesitancy.

---

# 1. Setup and preparation

```{r setup, message=FALSE, warning=FALSE}
source("../R/load_packages.R")
source("../R/config.R")
source("../R/modeling_utils.R")
source("../R/visualization_utils.R")

# Load partitioned data
model_split <- readRDS("../data/final/model_split.rds")
train_df    <- model_split$train
test_df     <- model_split$test

# Prepare data for modeling
final_data <- prepare_fold_data(train_df, test_df)
counts <- table(final_data$y_train)
wts <- ifelse(final_data$y_train == "No", sum(counts)/(2*counts["No"]), sum(counts)/(2*counts["Yes"]))
```

Quick load back of bootstrap
```{r}
#m2_stability <- read.xlsx("../data/final/model_comparison_BOOT10K_08selectfreqM1_95CIM2.xlsx", sheet = "M2_ALL_Boot_Predictors")
#m1_stability <- read.xlsx("../data/final/model_comparison_BOOT10K_08selectfreqM1_95CIM2.xlsx", sheet = "M1_ALL_Boot_Predictors")
```

# 2. Retraining the Baseline Model
We retrain the Elastic Net on the full 80% training set.
```{r}
set.seed(123)
m1_results <- fit_en_helper(final_data$x_train, final_data$y_train, wts)

# Evaluate on untouched hold-out set
final_probs <- as.numeric(predict(m1_results$model, newx = final_data$x_test, s = m1_results$params$lambda, type = "response"))
m1_metrics  <- calculate_metrics(final_data$y_test, final_probs, m1_results$model)

print("--- Final Hold-out Performance (Baseline) ---")
print(m1_metrics)
```

Bootstrapping for coefficient stability with 95% CI
```{r}
boot_N = 10000 # takes around 45-50 minutes
tic()
set.seed(123)
m1_stability <- fit_boot_stability(
  x = final_data$x_train, 
  y = final_data$y_train, 
  wts = wts, 
  alpha = m1_results$params$alpha, # Use the alpha found in Nested CV (0.8 so far)
  n_boot = boot_N 
)
toc()
```

Since the 95CI only showed 4 vars, changed to 0.80 selection frequency threshold. 
```{r}
stability_alpha <- 0.80 
m1_stable_vars <- m1_stability %>%
  filter(Selection_Freq > stability_alpha) %>%
  pull(Variable)

m1_stable_coefs <- m1_stability %>% 
  filter(Selection_Freq > stability_alpha) %>% 
  rename(Coefficient = Mean_Beta)

message(sprintf(">>> M1 Selection: %d variables identified as robustly associated.", length(m1_stable_vars)))

# Plot only the stable coefficients
fig_m1_coefs <- plot_model_coefs(m1_stable_coefs, paste0("Baseline, Bootstrap = ", boot_N, ", Selection freq >= ", stability_alpha)) 
fig_m1_coefs
```


# 3. Discovering Hierarchical Interactions
We use the glinternet algorithm to discover hierarchical interactions among our top predictors.
```{r}
# Discover interactions among stable variables (using glinternet)
x_train_filt <- final_data$x_train[, m1_stable_vars, drop = FALSE]
inter_list <- run_hierarchical_en(x_train_filt, final_data$y_train)

# Add interactions as features 
x_train_inter <- apply_top_interactions(final_data$x_train, inter_list)
x_test_inter  <- apply_top_interactions(final_data$x_test, inter_list)
print(inter_list)
```

# 4. Retraining a model with hierarchical interactions 
```{r}
set.seed(123)
m2_results <- fit_en_helper(x_train_inter, final_data$y_train, wts)
m2_probs   <- as.numeric(predict(m2_results$model, newx = x_test_inter, s = m2_results$params$lambda, type = "response"))
m2_metrics <- calculate_metrics(final_data$y_test, m2_probs, m2_results$model)

print("--- Model Comparison: Baseline vs Interactions ---")
comparison_table <- rbind(Baseline = m1_metrics, Interactions = m2_metrics)
print(comparison_table)
```


Bootstrapping for coefficient stability with 95% CI
```{r}
tic() # around 50-55 minutes
set.seed(123)
m2_stability <- fit_boot_stability(
  x = x_train_inter, 
  y = final_data$y_train, 
  wts = wts, 
  alpha = m2_results$params$alpha, # Use the alpha found in Nested CV (0 Ridge so far)
  n_boot = boot_N
)
toc()
```

Because M2 is Ridge (alpha = 0) by definition we cannot use selection frequency -> using only stable vars 95%CI
```{r}
toc()
# Filter only the Stable Predictors (CI doesn't cross zero)
m2_stable_vars <- m2_stability %>%
  filter(Stable == "Yes") %>%
  pull(Variable)
m2_stable_coefs <- m2_stability %>%
   filter(Stable == "Yes") %>%
  rename(Coefficient = Mean_Beta)

# m2_stable_coefs <- m2_stability %>% filter(Stable == "Yes") %>% rename(Coefficient = Mean_Beta)
# stability_alpha <- 0.80 
# m2_stable_vars <- m2_stability %>% 
#   filter(Selection_Freq > stability_alpha)  %>% 
#   pull(Variable)
# 
# m2_stable_coefs <- m2_stability %>% 
#   filter(Selection_Freq > stability_alpha) %>% 
#   rename(Coefficient = Mean_Beta)

message(sprintf(">>> M2 Selection: %d variables identified as robustly associated.", length(m2_stable_vars)))

# Plot only the stable coefficients
fig_m2_coefs <- plot_model_coefs(m2_stable_coefs, paste0("Interaction, Bootstrap = ", boot_N, " 95% CI")) 
fig_m2_coefs
```
Additional visualization for Bootstrapped results 
```{r}
# M1
fig_m1_robust <- plot_model_coefs_boot(m1_stability, "(Baseline Model)", boot_N)
print(fig_m1_robust)

# M2
fig_m2_robust <- plot_model_coefs_boot(m2_stability, "(Interaction Model)")
print(fig_m2_robust)
```

# Saving results 
```{r}
ggsave(file.path("../figures/", paste0("Main_M1_BOOT10K_08selectfreqM1_95CIM2.pdf")),fig_m1_coefs, width = 16, height = 15)
ggsave(file.path("../figures/", paste0("Main_M2_BOOT10K_08selectfreqM1_95CIM2.pdf")), fig_m2_coefs, width = 16, height = 15)
ggsave(file.path("../figures/", paste0("Main_M1_BOOT10K_08selectfreqM1_95CIM2_ROBUST.pdf")),fig_m1_robust, width = 16, height = 15)
ggsave(file.path("../figures/", paste0("Main_M2_BOOT10K_08selectfreqM1_95CIM2_ROBUST.pdf")), fig_m2_robust, width = 16, height = 15)

m_compare_table <- list(
  "Model_Comparison"   = comparison_table,
  "Hierarchical_Interactions_All" = inter_list, 
  "M1_Stable_Predictors"  = m1_stable_coefs,
  "M2_Stable_Predictors"  = m2_stable_coefs,
  "M1_ALL_Boot_Predictors"  = m1_stability,
  "M2_ALL_Boot_Predictors"  = m2_stability,
  "M1_All_Predictors" = get_top_predictors(m1_results$model, m1_results$params, 0),
  "M2_All_Predictors" = get_top_predictors(m2_results$model, m2_results$params, 0),
  
  "M1_Params" = data.frame(m1_results$params),
  "M2_Params" = data.frame(m2_results$params)
)
# Save everything into one multi-sheet Excel file
openxlsx::write.xlsx(m_compare_table,
                     file = "../data/final/model_comparison_BOOT10K_08selectfreqM1_95CIM2.xlsx")

```


# Post-hoc analysis 
```{r}
cluster_table <- list()
target_models <- c("m1", "m2")
for (target in target_models) {
  # Setup
  current_results <- get(paste0(target, "_results"))
  current_coefs   <- get(paste0(target, "_stable_coefs"))
  current_vars    <- get(paste0(target, "_stable_vars"))
  current_x_train <- if(target == "m2") x_train_inter else final_data$x_train
  current_x_test  <- if(target == "m2") x_test_inter else final_data$x_test
  
  # Interactions
  inter_pages <- list() 
  if (target == "m2" && any(grepl("int", current_coefs$Variable))){
    message(">>> Generating interaction plots for M2...")
  inter_plots <- plot_en_interactions(current_coefs, train_df, "take_vaccination", wts)
  inter_pages <- split(inter_plots, ceiling(seq_along(inter_plots) / 4)) %>%
    map(~wrap_plots(.x, ncol = 2, nrow = 2) + plot_annotation(title = "Interactions Analysis"))
  } 
  # Correlations
  p_cor_model <- plot_category_cor(data.frame(current_x_train), current_vars, 
                                   paste0(toupper(target), " Top Predictors"))
  
  # SHAP 
  shp_obj <- calculate_shaps(current_results, current_x_test)
  p_shap_summary <- plot_shap_summary(shp_obj, current_vars)
  
  shap_pdp_list <- map(current_vars, ~{
    shapviz::sv_dependence(shp_obj, v = .x, alpha = 0.5) + 
      fig_theme() + labs(title = make_pretty(.x))
  })
  shap_pages <- split(shap_pdp_list, ceiling(seq_along(shap_pdp_list) / 4)) %>%
    map(~wrap_plots(.x, ncol = 2, nrow = 2) + plot_annotation(title = "SHAP Dependency Profiles"))
  
  # Clustering
   vars_for_clustering <- current_vars
  if (target == "m2") {
    vars_for_clustering <- vars_for_clustering[!grepl("^int_", vars_for_clustering)]
  }
  
  clusters <- run_phenotype_clustering(current_x_train, final_data$y_train, vars_for_clustering, k = 3)
  p_clus_val  <- wrap_plots(clusters$validation_plots, ncol = 1) + plot_annotation(title = "Clustering Validation")
  p_clus_sig  <- plot_phenotype_profiles(clusters)
  p_clus_tsne <- plot_cluster_validation(clusters, paste0("(", toupper(target), ")"))
  cluster_summary <- clusters$data %>%
    group_by(Phenotype) %>%
    summarise(
      N_Total = n(),
      N_Hesitant = sum(Outcome == "No"),
      N_Intenders = sum(Outcome == "Yes"),
      Percent_Hesitant = round((N_Hesitant / N_Total) * 100, 1),
      Stability_Score = round(clusters$stability[first(as.numeric(Phenotype))], 3)
    )
  print(cluster_summary)
  cluster_table[[paste0(toupper(target), "_Clusters")]] <- cluster_summary
  
  # Combine 
  
  report_plots <- list(
    if (target == "m2") inter_pages else NULL, 
    list(p_cor_model, p_shap_summary),
    shap_pages,
    list(p_clus_val, p_clus_sig, p_clus_tsne)
  ) %>% purrr::flatten() %>%
    purrr::compact()
  
  # Save
  report_path <- file.path("../figures/", paste0("Interpretation_Report_", target, "_BOOT10K_08selectfreqM1_95CIM2.pdf"))
  save_report_pdf(report_plots, report_path, width = 16, height = 15)
  message(">>> FINISHED: ", report_path)
}

# Saving cluster table 
openxlsx::write.xlsx(cluster_table,
                     file = "../data/final/cluster_comparison_BOOT10K_08selectfreqM1_95CIM2.xlsx")

```


